---
bibliography: bibliography.bib
output:
  pdf_document: default
  html_document: default
  word_document:
    md_extensions: -autolink_bare_uris
---
<!--
%\VignetteEngine{knitr::rmarkdown}
%\VignetteIndexEntry{QUBIC vignette}
-->

QUBIC tutorial
==============

The `QUBIC` package is a biclustering package, with source code upgrading from C code to C++ code.
The updated source code can avoid memory allocation error and is much efficient than the original one[@li09].
Based on our preliminary analysis, it can save 40\% running time on a plant microarray data.

## Other languages

If R is not your thing, there is also a C version of [`QUBIC`](https://github.com/maqin2001/QUBIC).

## Help 

If you are having trouble with this R package, contact [the maintainer, Yu Zhang](mailto:zy26@jlu.edu.cn). 

## Install and load

Stable version from BioConductor


```r
source("https://bioconductor.org/biocLite.R")
biocLite("QUBIC")
```

<!-- Or development version from GitHub


```r
install.packages("devtools")
devtools::install_github("ropensci/alm")
```
-->
Load `QUBIC`


```{r, message=FALSE}
library("QUBIC")
```

## Random matrix with embedded bicluster


```{r}
test <- matrix(rnorm(5000),100,50)
test[11:20,11:20] <- rnorm(100,3,0.3)
res<-biclust(test, method = BCQU())
summary(res)
show(res)
names(attributes(res))
```

## References
